Stability of variational quantum dynamics with Gaussian approximants

### 2013-07-10 paper.tex
reorganisation

When the first quantum physicists understood that measurements are essentially uncertain, they naturally wondered what conditions would produce the normal law of probability, and discovered gaussian wave packets.  Once they got sick of the free particle and turned to the harmonic oscillator, they would have know to look for its ground state, and how it oscillated in the classical limit.  Gaussian wave packets were the answer to both questions.  Now, with their width fixed so that the action of a cycle was $\hbar$, they became coherent states.

When quantum mechanics was applied to more complicated systems, the equations of motion could not be solved in closed form.  Several approaches were used to approximate them: one of the most versatile was the variational method, in which a class of approximants was specified that could be treated exactly, and it was determined which of them most nearly solved the problem.  The most obvious application was in finding ground states, where the best solution was the one with the lowest energy.  Variational methods were also applied to quantum dynamics, by seeking approximants for which the difference between the two sides of Schrödinger's equation had the least norm in Hilbert space.  Gaussian wavepackets were an obvious choice for the class of approximants.  They are distinguished by their central position and momentum, and their scale: useful things to know about the solution to any problem.  Also, the gaussian packets and their derivatives are easy to manipulate algebraically.

@@

In the 1970s, Heller put the two ideas together, and proposed simulating quantum dynamics variationally, using superpositions of coherent states as the approximant.  This is an appealing approach.  Every quantum state can be represented arbitrarily well as the number of components in the superposition becomes large.  The experience of quantum optics shows that many systems can be approximated by a reasonable number of components, and variational dynamics should not have the covergence problems the Monte-Carlo methods do.  There have been several attempts to apply this numerically.  In all cases, numerical instability has prevented the superposition being expanded to the point of convergence.

@@

@@

@@

@@

Not a

Note that the gaussian entire functions are a vector subspace of the square integrable functions.

@@

@@

I've already figured out how to find the change in the condition number by pertubation theory.  That does require knowing the largest and smallest singular directions of $|Dψ〉$; perhaps there is a way to approximate them without finding an SVD.  How do I find the angle between $H|ψ〉$ and the space $|ψ〉$ can move in?  This seems to require shifting $z$ along each singular direction, solving the shifted least-squares problem, and comparing the residuals.  Is there a better way?

The accuracy of the pertubation method is shown below.

@@

@@

\item{2.} Measuring state differences in a truncated fock space makes the fit stable and accurate.  Of course, the truncated state is different from the original one, so the original state is not fit precisely.  When this was done, the least squares problem involved 6 fock coefficients, and the amplitude vector has 10 elements.  So almost all the truncated problems were underdetermined: some singular values of the Jacobian are exactly zero.  Matlab claims that backslash solves underdetermined problems by giving a solution with as many zero elements as possible.  The documentation doesn't say how it chooses which elements to make zero.  That's different from the other regularisation methods we've tried.  Of course, the problems are being solved inside Matlab's optimisation routine, and I don't know what it's doing.

\item{3.} The difference between the truncated state and the original one fails to satisfy the Picard condition when with respect to $|Dψ〉$.  What about $H|ψ〉$?

@@

A third method is the semi-implicit one, with the formulae
$$v^{\prime n}=vⁿ+½τf^{\prime n}\qquad v^{n+1}=vⁿ+τf^{\prime n}.$$
With the Hamiltonian, this becomes
$$c^{n+1}=cⁿ-iτH\biggl(1+{iτ\over 2}H\biggr)^{-1}cⁿ.$$
If Hilbert space is considered as a real vector space, complex multiples of the eigenstates of the Hamiltonian form planes instead of lines.  The eigenstates evolve as $e^{iωt}$, so, when the dynamics are projected onto an eigenplane, the state traces out a circle, with radius equal to the absolute value of the the coefficient of the eigenstate when the initial state is expanded.  Different eigenstates circle the origin with different frequencies.

The initial state was the ground state of the Morse oscillator potential, given various momentum kicks.  The number of components in the superposition representing this was varied.  The initial superposition was found by the AMPL optimisation program, with the constraint that no component would have a weight less that ?.

Coding this will be quite involved, and the result might not be very efficient.  However, if it allows physically chosen bases to be used in  variational simulations, it is likely to be worthwhile.

To test this idea, a five component superposition was fit to an amplitude 2 coherent state.  The Hilbert space residual was $1.3×10^{-3}$.  The amplitudes and norms of the components are shown above; the fitting routine constrained the norm of each component to be at least 0.1.  Note that components 1 and 3 ended up with very similar amplitudes.

Below is a plot of the condition number of $|Dψ(z)〉$ for this superposition, as the value of $z$ varies by $0.1$ in every direction.  The top line shows the effects of changing $φ$, the bottom line $α$, components are shown left to right.  Changes to real parts are black, imaginary parts red.  There are a few things to note.  The condition is very bad, with number $1.4×10^{11}$.  This is entirely due to the similar amplitudes of components 1 and 3: those graphs are on a different scale to the others, and it's logarithmic.  As they move closer and become identical, the condition number increases $10^{14}$.  No doubt it would go higher if we sampled the graphs more finely, but this is huge—reciprocal floating point epsilon is $5×10^{15}$.  If the separation between these components is doubled, the condition number improves by an order of magnitude.

The experiment with following the exact solution by nonlinear optimisation was repeated.  Instead of constraining the weights of the components, the logarithm of the condition number of the derivative was added in quadrature to the residual of the fit to give a cost.  This worked much better, getting stably to $t=0.4$ instead of under 0.1.  At that point, the residuals increased to order 1, and the superposition stopped following the exact solution.  This cold mean that 5 components weren't enough to represent those states, except that the same thing happened with 10 components.

Note that the optimisation is a nonlinear least squares problem, and special solvers exist for this case.

Some furt

>>>>

%% Background

When the first quantum physicists realised that the position of a particle is essentially uncertain, they were destined to seek the conditions that cause it to follow a normal law of error, and discover the gaussian wave packet.  They soon understood the free particle and turned to the harmonic oscillator, investigating its ground state and how its excited states generate sinusoidal motion in the classical limit.  These problems were also solved by the gaussian wave packet.  With its width fixed by equipartition of energy and the uncertainty principle, it became the coherent state.

The simplest mechanical systems allow Schrödinger's law of motion to be solved in closed form, but this luck doesn't last.  Most quantum problems must be solved approximately.  This can be done variationally, specifying a class of approximants that can be manipulated exactly, then finding one of them that most nearly solves the problem.  To test how well a putative solution approximates the dynamics of a system, it can be substituted into each side of Schrödinger's equation, and the difference between them evaluated by the usual norm in Hilbert space.  It remains to choose a class of approximants, and the gaussian wave packets are an common choice.  They differ in their scale and their central position and momentum, useful properties to determine for any quantum state.

@@

In the 1970s, Heller put the two ideas together, and proposed simulating quantum dynamics variationally, using superpositions of coherent states as the approximant.  This is an appealing approach.  Every quantum state can be approximated arbitrarily well as the number of components in the superposition becomes large.  The experience of quantum optics shows that many systems can be approximated by a reasonable number of components, and variational dynamics should not have the covergence problems the Monte-Carlo methods do.

%% Problem

Since Heller suggested it, there have been several attempts to apply this numerically.  In all cases, numerical instability has prevented the superposition being expanded to the point of convergence.
%% Consequence
These attempts have aimed to solve chemistry problems, involving dozens of atoms, where the lack of convergence limited the precision that could be obtained, but superpositions with a few components, and ad-hoc methods to improve the stability, still gave useful results.  The kind of problems that people approach with phase space methods come from atom optics or condensed matter; they involve thousands or millions of atoms, and the convergence problems would prevent any useful information being obtained.  A systematic approach to regularisation is required in order to approach these problems variationally.  The prospective rewards are great: these problems include high temperature superconductivity and the dynamics of the early universe, topics on which today's physicists expend massive efforts.

@@

%% Phase space expansion as a linear operator

%% SVE lemma

@@

The norm of $Gg$ is 1, so this suggests the singular values of $G$ are $π^{-½}$.

%% Consequences of lemma

@@

%% Discrete phase space

The obvious discrete version of the problem, where the state is approximated by a finite linear combination of coherent states, doesn't relate to the continuous problem in an obvious way.  The weights of the components don't simple sample $f(α)$: they also adjust for how close the sample points are, and how much the $|α_i〉$ overlap.  The operator $G$ is probably being disceretised by some kind of Galerkin formula; I don't know what the basis functions are in phase space and Hilbert space.

It would be useful to find a quadrature rule, with weights $w_i$, such that 
$$∑_i w_if(α_i)|α_i〉≈\int f(α)|α〉d²α.$$
Suppose we aim to be exactly right, in the case that $f(α)$ is the normalised entire function representing the left hand side, i.e.,
$$f(β)=∑_i w_if(α_i)〈β|α_i〉.$$
(I'm assuming we put the factor $1/π$ in $G$.)  This doesn't have a solution for general superpositions of coherent states.  Surely there's some quadrature rule for a sampled phase-space function?

Using the Galerkin criterion, there doesn't seem to be one where the coefficient of $|α_i〉$ only depends on $f(α_i)$.  However, the Galerkin formula $〈A|A〉f=〈A|ψ〉$, where $〈α_i|ψ〉=f(α_i)$, seems like a good candidate for a more general rule.

%% Truncated Hilbert space

%% Why grids don't work, and variation necessary

@@

%% Fock space is cheating

%% Singular value decompositions

%% Picard plots

@@

%% Adapting the sample points to maintain condition

@@

I've figured out how to find the change in the condition number by pertubation theory.  That does require knowing the largest and smallest singular directions of $|Dψ〉$; perhaps there is a way to approximate them without finding an SVD.  How do I find the angle between $H|ψ〉$ and the space $|ψ〉$ can move in?  This seems to require shifting $z$ along each singular direction, solving the shifted least-squares problem, and comparing the residuals.  Is there a better way?

The accuracy of the pertubation method is shown below.  The state is coherent, with ampltiude 2.  It was sampled at 5 amplitudes, shown on the left hand plot.  The weights were assigned by inverting the discretised $G$ operator on these sample points, with small singular values truncated by Matlab's {\tt pinv} function.  The the moduli of the resulting weights are shown in the right hand plot.  This method of fitting usually gives a Hilbert space residual around 0.01.

The lower plot shows how the smallest singular value of the discrete $G$ changes, as the sample points are shifted.  The black lines show changes in the real parts of the amplitudes, the red lines in the imaginary parts.  The dotted black line shows the derivatives for the real parts calculated by pertubation theory, as follows.  The derivatives of the matrix $〈N|G|A〉$  with respect to each real part were computed by symmetric difference, with $h=0.01$.  Here, $|Α〉=(|α₁〉 |α₂〉 \cdots |α_R〉)$.  A matrix 
$$M=\pmatrix{0 & 〈N|G|A〉\cr 〈A|G|N〉 & 0}$$
was constructed; by substituting $〈N|G|A〉=USV†$, it is clear that the vector
$$x={1\over √2}\pmatrix{u_R \cr v_R}$$
is an eigenvector, with eigenvalue $σ_R$, and norm 1.  (Provided $N+1>R$.)  From first-order pertubation theory, the derivative of this eigenvalue with respect to amplitude $i$ is
$$∂_i σ_i=x†∂_iMx.$$
As shown in the plots, lines of this slope fit the singular value very well.

@@

Below is a plot of the condition number of $|Dψ(z)〉$ for a superposition, as the value of $z$ varies by $0.1$ in every direction.  The top line shows the effects of changing $φ$, the bottom line $α$, components are shown left to right.  Changes to real parts are black, imaginary parts red.  There are a few things to note.  The condition is very bad, with number $1.4×10^{11}$.  This is entirely due to the similar amplitudes of components 1 and 3: those graphs are on a different scale to the others, and it's logarithmic.  As they move closer and become identical, the condition number increases $10^{14}$.  No doubt it would go higher if we sampled the graphs more finely, but this is huge—reciprocal floating point epsilon is $5×10^{15}$.  If the separation between these components is doubled, the condition number improves by an order of magnitude.

%% Shape of Hilbert space trajectories

If Hilbert space is considered as a real vector space, complex multiples of the eigenstates of the Hamiltonian form planes instead of lines.  The eigenstates evolve as $e^{iωt}$, so, when the dynamics are projected onto an eigenplane, the state traces out a circle, with radius equal to the absolute value of the the coefficient of the eigenstate when the initial state is expanded.  Different eigenstates circle the origin with different frequencies.

%% Discrete time and stiffness

A third method is the semi-implicit one, with the formulae
$$v^{\prime n}=vⁿ+½τf^{\prime n}\qquad v^{n+1}=vⁿ+τf^{\prime n}.$$
With the Hamiltonian, this becomes
$$c^{n+1}=cⁿ-iτH\biggl(1+{iτ\over 2}H\biggr)^{-1}cⁿ.$$

@@

Measuring state differences in a truncated fock space makes the fit stable and accurate.  Of course, the truncated state is different from the original one, so the original state is not fit precisely.  When this was done, the least squares problem involved 6 fock coefficients, and the amplitude vector has 10 elements.  So almost all the truncated problems were underdetermined: some singular values of the Jacobian are exactly zero.  Matlab claims that backslash solves underdetermined problems by giving a solution with as many zero elements as possible.  The documentation doesn't say how it chooses which elements to make zero.  That's different from the other regularisation methods we've tried.  Of course, the problems are being solved inside Matlab's optimisation routine, and I don't know what it's doing.

%% The complete algorithm

%% Numerical experiments

%%%%

@@
